Circular Distributional Trees and Forests with an Application to Wind Direction Forecasting

### 2019-11-21 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Remove all comments and clean up Rnw file
- additionally spell check
- Add draft version 2019-11-20
- Update makefile (for diff)


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2894 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f82750fb5fb8c4263ddf38abe3b4a195471ef852

To evaluate the proposed circular trees and forests, a case study on probabilistic 
forecasting of wind direction for two airports in Austria is presented. 
In addition to the tree-based models, alternative common approaches are 
used as a benchmark in order to assess the predictive performances of circular trees 
and forests.

%As circular data can be found in many different areas, the development of
%appropriate statistical models accounting, i.a., for periodicity has become a
%challenging and important research field. For probabilistic forecasting
%typically the von Mises distribution is employed in an additive model
%framework. Combining their underlying distributional setup with tree-based
%estimation methods, circular distributional trees will be presented. The resulting
%distributional trees are easy to interpret, can detect non-additive effects,
%and select covariates and their interactions automatically. By combining
%ensembles of circular trees a stabilized circular forest can build, which
%automatically regularizes the estimation and, hence, allows for modeling smooth effects.

%To validate the proposed circular trees and forests, a case study on
%probabilistic forecasting of wind direction will be shown for two airports in
%Austria. In addition to the tree-based models, alternative commonly used
%approaches will be employed to compare their predictive performances to the
%proposed circular trees and forest.

%%%%%%

%As circular data can be found in many different subject areas accounting for periodicity has 
%become an important aspect in establishing appropriate statistical models.
%For probabilistic modeling of circular data the von Mises distribution is widely used.
%To capture how its parameters change with covariates, a distributional regression tree model
%is proposed as an alternative to more commonly-used additive models. The resulting
%distributional trees are easy to interpret, can detect non-additive effects, and select
%covariates and their interactions automatically. In order to stabilize and regularize
%the model an ensemble of distributional trees is combined yielding a distributional
%forest which allows for modeling smooth effects as well.
%For illustration, hourly wind direction forecasts are obtained at two airports in Austria
%(Vienna and Innsbruck) based on a set of meteorological measurements from each of the 
%airports and surrounding weather stations.
}
\Keywords{Distributional Trees and Forests; Circular Response; von Mises Distribution}

@@

\begin{document}

\section{Introduction}
\label{sec:introduction}

Circular data can be found in a variety of applications and subject areas, 
e.g., hourly crime rate in the social-economics, animal movement direction or 
gene-structure in biology, and wind direction as one of the most important weather 
variables in meteorology. Fitting a statistical model to this type of data requires
the incorporation of its specific feature of periodicity, i.e., in case of angular
data the fact that values are restricted to an interval such as $[0,2\pi)$ with
$0$ being equivalent to $2\pi$.

In many approaches to model circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution which is also known 
as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
This is also the basis which early regression models for circular data are built on as they 
regress the parameter(s) of the von Mises distribution on covariates.
One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where the circular mean, i.e., the location parameter of 
the von Mises distribtuion is predicted by a linear combination of covariates.
\cite{Johnson+Wehrly:1978} took this idea one step further by plugging in a link function such that the values of the 
linear predictor are transformed to a restricted interval of length $2\pi$. 
This generalized linear model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
introducing another specific transformation function which maps the values of the real line to the 
unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter of the von Mises distribution, to be modeled depending
on covariates, not only in separate models but also in a mixed model. 
%There, within an iterative 
%optimization algorithm the two parameters are estimated and improved stepwise until convergence. 
%To obtain proper initial values it is recommended to apply two separate models where either one of 
%the parameters is fixed while the other is predicted employing covariates in a GLM-type model. 
While being built on well elaborated theory this methodology 
%and the contained optimization problem have 
has shown to be very challenging in their application, mainly due to the
complexity encountered in optimizing the log-likelihood function which is not
globally concave.  Therefore, highly informative starting values are crucial
for the model to converge at all \citep{Pewsey+Neuhaeuser+Ruxton:2013,
Gill+Hangartner:2010}.  In order to avoid this strong dependence on appropriate
initial values, \cite{Mulder+Klugkist:2017} presented an alternative approach
to regression modeling of circular data by employing a GLM in a Bayesian
framework.  The listed methods and most of their further developments
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive
regression models.  While these generally provide a powerful toolbox for modeling
particularly smooth effects, they may not be able to capture all types of
effects adequately.  Especially abrupt changes might not be detected properly
owing to the enforced additive structure and model assumptions, such as a
possibly highly non-linear link function and the form of linear combinations.

@@

Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the 
empirical (gray) and fitted von Mises (red) distribution of wind direction in each terminal 
node. Based on the fitted location parameters $\hat \mu$, the subgroups can be distinguished 
into the following wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
east to west, nodes 4 and 5). 
(2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
Innsbruck (from south-east to north-west, node 8). 
(3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
nodes 10, 12 and 13). 
Node~7 captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low estimated 
concentration $\hat \kappa$. In terms of covariates, the lagged wind direction 
(``persistence'') is mostly responsible for distinguishing the broad wind regimes 
listed above while the pressure gradients and wind speed separate between subgroups 
with high vs.\ low precision. Note that in the meteorological context wind direction 
is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).

\begin{figure}[t]
\centering
\includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
\caption{Fitted tree based on the von Mises distribution for 1-hourly wind direction forecasts 
at the airport of Innsbruck.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates
employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
and pressure gradients ($\text{dpressure; hPa}$) west, east and south of the airport,
all lagged by one hour.}
\label{fig:tree_ibk} 
\end{figure}

@@

\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper circular
probability distribution, i.e., a probability distribution which accounts for the periodicity
of circular data.
Generally, this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
to use \citep{Fisher:1993} is provided by the von Mises distribution,
a purely circular distribution which is also known as ``the circular normal distribution''
and commonly applied for probabilistic modeling of circular data.
Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration parameter $\kappa > 0$ 
the density of the von Mises distribution for an observation $y \in (-\pi, \pi]$ is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.

The corresponding log-likelihood function is defined by
\begin{equation}
\begin{aligned}
\ell(\mu, \kappa; y) &= \log(f_\mathrm{vM}(y;\mu, \kappa)) \\
&= -\log(2 \pi I_0(\kappa)) +  \kappa \cos(y - \mu).\label{equ:vM:loglik}
\end{aligned}
\end{equation}

@@

By means of the score function
\begin{equation}
\label{eq:scores}
\begin{aligned}
s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
\smallskip &,
\quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y_i) \bigg) 
\\
&= \bigg(\kappa \sin(y-\mu)
\smallskip &,
\quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
\end{aligned}
\end{equation}
an indicator of deviation and in that way a measure of goodness of fit of the model can be obtained 
for each observation $y_i$ and fitted parameter. Moreover, the optimization problem in Equation~\ref{eq:optim} can also 
be defined by 
%setting the sum of the first derivative of the log-likelihood function evaluated for all observations to zero, i.e., by setting
\begin{equation}
\sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}

\begin{figure}[t]
%\centering
%\hspace{-1cm}
\begin{minipage}{.05\textwidth}
\hfill
\end{minipage}%
\begin{minipage}{.45\textwidth}
  %\centering
  \vspace{-1em}
  \includegraphics[width=1.05\linewidth]{density_linear.pdf}
  \end{minipage}%
\begin{minipage}{.45\textwidth}
  %\centering
  \includegraphics[width=1.2\linewidth]{density_circular.pdf}
\end{minipage}
\begin{minipage}{.05\textwidth}
\hfill
\end{minipage}%
\caption{Illustration of a distributional model fitted to circular data by 
  maximum likelihood estimation employing the von Mises distribution. 
  Left panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. 
  Right panel: circular illustration.
  In both panels the empirical histogram (gray) and fitted density (red line) are depicted along 
  with the estimated location parameter (red hand).
  }
  \label{fig:densities} 
  %\includegraphics[width=4cm]{density_vM.pdf}
\end{figure}

Figure~\ref{fig:densities} depicts a distributional model fitted to circular data
employing the von Mises distribution. While the left panel only serves as an illustration 
of the corresponding  ``linear'' description restricted to an interval of length $2\pi$ but 
consequently not accounting for periodicity, the right panel displays an appropriate 
representation of a circular quantity on the unit circle. 
% on observed wind direction data from the airport of Innsbruck.
Looking at the fitted density (red line) and estimated location parameter (red hand) this example 
shows that reasonably well fitting probabilistic models can be established for circular 
data by specifying an appropriate distribution family such as the von Mises distribution
and estimating its parameters via maximum likelihood. 
However, this procedure considers only the response variable. Including covariates 
can improve any model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types of data,
including circular data.

\section{Circular distributional trees and forests}
\label{sec:tree_forest}
A common way of including covariates is to regress the model parameters on covariates.
In the case of a probabilistic model for a circular response this corresponds to expressing the 
location parameter $\mu$ and the concentration parameter $\kappa$ of the von Mises distribution
by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
appropriate link function plugged in as in GLMs. This kind of approach is presented for example
in the mixed model of \cite{Fisher+Lee:1992} and \cite{Fisher:1993}. 
%While applying the mixed model 
%requires expert user knowledge due to its complexity as discussed in Section~\ref{sec:introduction},
%\cite{Mulder+Klugkist:2017} tried to facilitate applications by their Bayesian regression model for
%a circular response variable.
Apart from the methodology developed specifically for circular data more general approaches to 
regressing model parameters on covariates are commonly used such as for example generalized additive 
models for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} 
where each distribution parameter is estimated by its own generalized additive model (GAM). 
%However, an adaption of this methodology to circular data can be very complex due to a
%challenging representation of smooth effects on the unit circle.
%However, an adaption of this methodology to circular data can not be obtained by the straight 
%forward approach of simply wrapping the modeled additive effects
%around the unit circle since the parameters of the von Mises distribution and their 
%dependence/interaction differs strongly from the linear case and therefore requires specific handling.
However, difficulties might arise in case of non-additive changes and a proper model 
specification can be very challenging, particularly for a high number of covariates and no information
on possible interactions.
As an alternative regression approach which can deal with these problems, we introduce a circular probabilistic tree-structured model in this section.

@@

In the applied tree-building algorithm, in each node the crucial decision on how and where to split the data is based on model
scores which are obtained by evaluating the score function $s$ 
for the individual observations and parameter estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its two distribution parameters and a data set 
of $n$ observations this yields an $n \times 2$ matrix that can be employed as a discrepancy measure,
capturing how well each given observation conforms with the estimated location $\hat{\mu}$ and precision
$\hat{\kappa}$, respectively.
To capture dependence on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a permutation test~(CTree).
By doing so in each partitioning step, the covariate with the highest significant association (i.e., lowest
significant $p$-value, if any) is selected for splitting the data. The corresponding split point
is chosen either by optimizing the log-likelihood~(MOB) or a two-sample test statistic~(CTree)
over all possible partitions.
%and the covariate corresponding to the strongest change is selected as split varialbe.
%Next a split point is chosen within the selected covariate by maximizing a partitioned likelihood.
This procedure is repeated recursively until there are no significant parameter instabilities or until another stopping criterion is met (e.g., subgroup size or tree depth).

%In particular, starting with the full learning sample, the steps for building a circular 
%distributional tree are:
%\begin{enumerate}
%\item Estimate the parameter pair $(\hat{\mu}, \hat{\kappa})$ via maximum likelihood for the observations
%  in the current (sub)sample.
%\item Test for associations (or instabilities) of the scores $s(\hat{\mu}, \hat{\kappa}, y_i)$ 
%  and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
%\item Split the sample along the partitioning variable $Z_l^*$ with the
%  strongest association or instability. Choose the breakpoint leading to the highest
%  improvement in model fit.
%  %with the highest improvement in the log-likelihood or the highest discrepancy.
%\item Repeat steps 1--3 recursively in the subsamples until these become too
%  small or there is no significant association/instability (or some other
%  stopping criterion is reached).
%\end{enumerate}

%To test for associations between scores and covariates in step~3 different strategies can be applied
%such as permutation tests introduced by \cite{Hothorn+Hornik+VanDeWiel:2006} or asymptotic M-fluctuation
%tests for parameter instability \citep{Zeileis+Hornik:2007}. In order to select the covariate with the %highest significant association the one covariate that corresponds to the lowest significant $p$-value (if %any) is chosen for splitting the data. Within the range of this split variable the split point leading to %the highest improvement in model fit is chosen either by optimizing the log-likelihood \citep[as in the %MOB algorithm,][]{Zeileis+Hothorn+Hornik:2008} or a two-sample test statistic  over all possible partition%s \citep[as in the CTree algorithm][]{Hothorn+Hornik+Zeileis:2006}.

Once a distributional tree model is fitted it can be employed to obtain probabilistic predictions
for a possibly new set of observed covariates $\bm{z} = (z_1, \ldots, z_m)$.
Starting at the root node the tree structure leads the observation to a terminal node, depending on the values of the covariates, where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for
the corresponding subset of learning observations. 
This can also be expressed by employing weights which indicate whether the $i$-th learning observation and the observation $\bm{z}$ belong to the same terminal
node:
\begin{equation}
w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)),
\end{equation}
where $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th out of $B$ segments
partitioning the covariate space in disjoint subsets. 
In that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted
von Mises distribution for a given $\bm{z}$ is obtained by
\begin{equation}
(\hat{\mu},\hat{\kappa})(\bm{z}) = \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot \ell(\mu,\kappa; y_i).
\end{equation}

Therefore, the same parameter pair is estimated for all observations
belonging to the same terminal node. This facilitates the application as the parameter estimates do 
not need to be recalculated for each (new) observation via maximum likelihood but can be extracted 
directly from the learning sample and the fitted model.

@@

\subsection{Circular forests}
\label{sec:circforest}
A natural extension of circular distributional trees are ensembles or forests 
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by \cite{Breiman:2001} average the predictions of an ensemble
of trees, each built on a subsample or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions by adaptive local 
likelihood estimation of the distributional parameters 
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor''
weights $w_i^{forest}(\bm{z})$ 
%\citep{Lin+Jeon:2006}
is obtained that is based on the number of trees in which $\bm{z}$ is assigned to the same terminal 
node as each learning observation $y_i, i \in \{1,\ldots,n\}$.
Hence, for a forest of $T$ trees the weights are calculated via
\begin{equation}
w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
\frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree.
Therefore, similar observations ending up more often in the same terminal node have higher weights
and in that way a stronger influence in the estimation process.

In that way a specific set of weights can be calculated for each observation yielding
its specific parameter estimates for the von Mises distribution
\begin{equation}
(\hat{\mu},\hat{\kappa})(\bm{z}) = \operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i^{forest}(z) \cdot \ell(\mu, \kappa; y_i). 
\end{equation}

@@

\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}
Accurate forecasts of wind directions are of great importance for risk
management in various fields such as agriculture, energy production or air safety control. 
For example, in order to direct airplanes to a safe landing, precise knowledge of wind 
direction for the next hour(s) at the respective airport is highly desirable and adequate 
prediction methods are required. In this section, a comprehensive case study on probabilistic
wind direction forecasting is provided for two airports in Austria with fundamentally
different site characteristics: Innsbruck with a rather unusual location for an
airport, located at the bottom of a deep Alpine valley, and Vienna located
within flat terrain in the East of Austria. In addition to the newly
introduced circular distributional trees and forests, the case study employs various
alternative approaches to forecast wind direction as a benchmark. The study is based on
$1$\,h and $3$\,h forecasts employing lagged observations in the vicinity of
the airports as predictor variables.

>>>>

To evaluate the proposed circular trees and forests, a case study on
probabilistic forecasting of wind direction for two airports in Austria is
presented.  In addition to the tree-based models, alternative common approaches
are used as a benchmark in order to assess the predictive performances of
circular trees and forests. 
}
\Keywords{Distributional Trees and Forests; Circular Response; von Mises Distribution}

@@

% -------------------------------------------------------------------
% Introduction
% -------------------------------------------------------------------
\begin{document}
\section{Introduction}
\label{sec:introduction}
Circular data can be found in a variety of applications and subject areas,
e.g., hourly crime rate in the social-economics, animal movement direction or
gene-structure in biology, and wind direction as one of the most important
weather variables in meteorology. Fitting a statistical model to this type of
data requires the incorporation of its specific feature of periodicity, i.e.,
in case of angular data the fact that values are restricted to an interval such
as $[0,2\pi)$ with $0$ being equivalent to $2\pi$.

In many approaches to model circular data it is assumed that the circular
variable of interest follows a circular distribution, in particular the von
Mises distribution which is also known as the ``circular normal distribution''
(see Section~\ref{sec:prob_circ} for more details). This is also the basis
which early regression models for circular data are built on as they regress
the parameter(s) of the von Mises distribution on covariates. One of the first
regression models with a circular response variable and linear covariates was
presented by \cite{Gould:1969} where the circular mean, i.e., the location
parameter of the von Mises distribution is predicted by a linear combination of
covariates. \cite{Johnson+Wehrly:1978} took this idea one step further by
plugging in a link function such that the values of the linear predictor are
transformed to a restricted interval of length $2\pi$. This generalized linear
model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
subsequently by \cite{Fisher:1993} introducing another specific transformation
function which maps the values of the real line to the unit circle. Moreover,
the presented methodology allows for both distribution parameters, the mean
direction and the concentration parameter of the von Mises distribution, to be
modeled depending on covariates, not only in separate models but also in a
mixed model. While being built on well elaborated theory this methodology has
shown to be very challenging in their application, mainly due to the complexity
encountered in optimizing the log-likelihood function which is not globally
concave. Therefore, highly informative starting values are crucial for the
model to converge at all \citep{Pewsey+Neuhaeuser+Ruxton:2013,
Gill+Hangartner:2010}. In order to avoid this strong dependence on appropriate
initial values, \cite{Mulder+Klugkist:2017} presented an alternative approach
to regression modeling of circular data by employing a GLM in a Bayesian
framework. The listed methods and most of their further developments
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive
regression models. While these generally provide a powerful toolbox for
modeling particularly smooth effects, they may not be able to capture all types
of effects adequately. Especially abrupt changes might not be detected
properly owing to the enforced additive structure and model assumptions, such
as a possibly highly non-linear link function and the form of linear
combinations.

@@

Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
empirical (gray) and fitted von Mises (red) distribution of wind direction in
each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes 4 and 5). (2) Downslope winds blowing across the Alpine crest
along the intersecting valley towards Innsbruck (from south-east to north-west,
node 8). (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be distinguished clearly into wind regimes
and consequently are associated with a very low estimated concentration $\hat
\kappa$. In terms of covariates, the lagged wind direction (``persistence'') is
mostly responsible for distinguishing the broad wind regimes listed above while
the pressure gradients and wind speed separate between subgroups with high vs.\
low precision. Note that in the meteorological context wind direction 
is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
%%
\begin{figure}[t]
  \centering
  \includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
  \caption{Fitted tree based on the von Mises distribution for 1-hourly wind
  direction forecasts at the airport of Innsbruck. In each terminal node the
  empirical histogram (gray) and fitted density (red line) are depicted along
  with the estimated location parameter (red hand). The covariates employed are
  wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
  ($\text{dpressure; hPa}$) west, east and south of the airport, all lagged by
  one hour.}
  \label{fig:tree_ibk} 
\end{figure}
%%

@@

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and commonly applied for probabilistic modeling of circular
data. Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in (-\pi, \pi]$ is given by:
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ \citep[see, e.g.,][for a more detailed
overview]{Jammalamadaka+Sengupta:2001}.

The corresponding log-likelihood function is defined by
%%
\begin{equation}
  \begin{aligned}
    \ell(\mu, \kappa; y) &= \log(f_\mathrm{vM}(y;\mu, \kappa)) \\
    &= -\log(2 \pi I_0(\kappa)) + \kappa \cos(y - \mu).\label{equ:vM:loglik}
  \end{aligned}
\end{equation}
%%

@@

By means of the score function
%%
\begin{equation}
  \label{eq:scores}
  \begin{aligned}
    s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
    \smallskip &,
    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y_i) \bigg) 
    \\
    &= \bigg(\kappa \sin(y-\mu)
    \smallskip &,
    \quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
  \end{aligned}
\end{equation}
%%
an indicator of deviation and in that way a measure of goodness of fit of the
model can be obtained for each observation $y_i$ and fitted parameter.
Moreover, the optimization problem in Equation~\ref{eq:optim} can also be
defined by 
%%
\begin{equation}
  \sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}
%%
%%
\begin{figure}[t]
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \vspace{-1em}
    \includegraphics[width=1.05\linewidth]{density_linear.pdf}
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \includegraphics[width=1.2\linewidth]{density_circular.pdf}
  \end{minipage}
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \caption{Illustration of a distributional model fitted to circular data by
    maximum likelihood estimation employing the von Mises distribution. Left
    panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. Right
    panel: circular illustration. In both panels the empirical histogram (gray)
    and fitted density (red line) are depicted along with the estimated location
    parameter (red hand).}
  \label{fig:densities} 
\end{figure}
%%

Figure~\ref{fig:densities} depicts a distributional model fitted to circular
data employing the von Mises distribution. While the left panel only serves as
an illustration of the corresponding ``linear'' description restricted to an
interval of length $2\pi$ but consequently not accounting for periodicity, the
right panel displays an appropriate representation of a circular quantity on
the unit circle. Looking at the fitted density (red line) and estimated
location parameter (red hand) this example shows that reasonably well fitting
probabilistic models can be established for circular data by specifying an
appropriate distribution family such as the von Mises distribution and
estimating its parameters via maximum likelihood. However, this procedure
considers only the response variable. Including covariates can improve any
model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types
of data, including circular data.

% -------------------------------------------------------------------
% Circular distributional trees and forests
% -------------------------------------------------------------------
\section{Circular distributional trees and forests}
\label{sec:tree_forest}
A common way of including covariates is to regress the model parameters on
covariates. In the case of a probabilistic model for a circular response this
corresponds to expressing the location parameter $\mu$ and the concentration
parameter $\kappa$ of the von Mises distribution by means of the available
covariates, e.g., in the form of a linear combination, possibly with an
appropriate link function plugged in as in GLMs. This kind of approach is
presented for example in the mixed model of \cite{Fisher+Lee:1992} and
\cite{Fisher:1993}. Apart from the methodology developed specifically for
circular data more general approaches to regressing model parameters on
covariates are commonly used such as for example generalized additive models
for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}
where each distribution parameter is estimated by its own generalized additive
model (GAM). However, difficulties might arise in case of non-additive changes
and a proper model specification can be very challenging, particularly for a
high number of covariates and no information on possible interactions. As an
alternative regression approach which can deal with these problems, we
introduce a circular probabilistic tree-structured model in this section.

@@

In the applied tree-building algorithm, in each node the crucial decision on
how and where to split the data is based on model scores which are obtained by
evaluating the score function $s$ for the individual observations and parameter
estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
two distribution parameters and a data set of $n$ observations this yields an
$n \times 2$ matrix that can be employed as a discrepancy measure, capturing
how well each given observation conforms with the estimated location
$\hat{\mu}$ and precision $\hat{\kappa}$, respectively. To capture dependence
on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a
permutation test~(CTree). By doing so in each partitioning step, the covariate
with the highest significant association (i.e., lowest significant $p$-value,
if any) is selected for splitting the data. The corresponding split point is
chosen either by optimizing the log-likelihood~(MOB) or a two-sample test
statistic~(CTree) over all possible partitions. This procedure is repeated
recursively until there are no significant parameter instabilities or until
another stopping criterion is met (e.g., subgroup size or tree depth).

Once a distributional tree model is fitted it can be employed to obtain
probabilistic predictions for a possibly new set of observed covariates $\bm{z}
= (z_1, \ldots, z_m)$. Starting at the root node the tree structure leads the
observation to a terminal node, depending on the values of the covariates,
where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for the
corresponding subset of learning observations. This can also be expressed by
employing weights which indicate whether the $i$-th learning observation and
the observation $\bm{z}$ belong to the same terminal node: \begin{equation}
w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in
\mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)), \end{equation} where
$\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th
out of $B$ segments partitioning the covariate space in disjoint subsets. In
that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$
specifying the predicted von Mises distribution for a given $\bm{z}$ is
obtained by 
%%
\begin{equation} 
  (\hat{\mu},\hat{\kappa})(\bm{z}) =
  \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot
  \ell(\mu,\kappa; y_i).
\end{equation}
%%
Therefore, the same parameter pair is estimated for all observations belonging
to the same terminal node. This facilitates the application as the parameter
estimates do not need to be recalculated for each (new) observation via maximum
likelihood but can be extracted directly from the learning sample and the
fitted model.

@@

\subsection{Circular forests}
\label{sec:circforest} 
A natural extension of circular distributional trees are ensembles or forests
that can improve forecasts by regularizing and stabilizing the model. Random
forests introduced by \cite{Breiman:2001} average the predictions of an
ensemble of trees, each built on a subsample or bootstrap of the original data.
A generalization of this strategy is to obtain weighted predictions by adaptive
local likelihood estimation of the distributional parameters
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019,
Hothorn+Zeileis:2017}. More specifically, for each possibly new
observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
$w_i^{forest}(\bm{z})$ is obtained that is based on the number of trees in
which $\bm{z}$ is assigned to the same terminal node as each learning
observation $y_i, i \in \{1,\ldots,n\}$. Hence, for a forest of $T$ trees the
weights are calculated via
%%
\begin{equation}
  w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
  \frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
%%
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree. Therefore, similar observations ending up more
often in the same terminal node have higher weights and in that way a stronger
influence in the estimation process.

In that way a specific set of weights can be calculated for each observation
yielding its specific parameter estimates for the von Mises distribution
%%
\begin{equation}
  (\hat{\mu},\hat{\kappa})(\bm{z}) = \operatorname{argmax}\displaylimits_{\mu,
  \kappa} \sum_{i=1}^n w_i^{forest}(z) \cdot \ell(\mu, \kappa; y_i). 
\end{equation}
%%

@@

% -------------------------------------------------------------------
% Case study
% -------------------------------------------------------------------
\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}
Accurate forecasts of wind directions are of great importance for risk
management in various fields such as agriculture, energy production or air
safety control. For example, in order to direct airplanes to a safe landing,
precise knowledge of wind direction for the next hour(s) at the respective
airport is highly desirable and adequate prediction methods are required. In
this section, a comprehensive case study on probabilistic wind direction
forecasting is provided for two airports in Austria with fundamentally
different site characteristics: Innsbruck with a rather unusual location for an
airport, located at the bottom of a deep Alpine valley, and Vienna located
within flat terrain in the East of Austria. In addition to the newly introduced
circular distributional trees and forests, the case study employs various
alternative approaches to forecast wind direction as a benchmark. The study is
based on $1$\,h and $3$\,h forecasts employing lagged observations in the
vicinity of the airports as predictor variables.
